FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2025-01-02 manuscript.Rmd
add svg code for DOT graphs

The process described here can be largely automated in R using the `theorytools` package; see the package vignette on FAIR theory, `vignette("fairtheory", package = "theorytools")`.
<!-- R-users can use the following code (see the package documentation for more information): -->

<!-- ``` -->
<!-- install.packages("theorytools") -->
<!-- library(theorytools) -->
<!-- # Use worcs to check if GitHub permissions are set: -->
<!-- library(worcs) -->
<!-- check_github() -->
<!-- # Create the theory repository: -->
<!-- fair_theory(path = "directory/for/theory", -->
<!--             title = "The Theory's Title", -->
<!--             theory_file = "path_to_existing_theory.txt", -->
<!--             remote_repo = "github_repo_name", -->
<!--             add_license = "cc0") -->
<!-- ``` -->

@@

@@

## Comparing Implementations of a Theory

As several authors have taken inspiration from the work by De Groot  [@degrootMethodologieGrondslagenVan1961],  
we compare our interpretation of the original theory to the interpretation of others.

Subsequently, Wagenmakers and colleagues modified the theory by *"[adding the] Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*:

@@

  subgraph cluster_discovery {
    label="Discovery";
    hypothesis [label="New hypothesis"];
    prediction [label="New prediction"];
  }
  data  [label="Old knowledge and old data"];      
  subgraph cluster_justification {
    label="Justification";
    test [label="Test on new data"];
    evaluation;
  }

  data -> hypothesis [label="Speculate & explore"];
  hypothesis -> prediction  [label="Deduce"];
  prediction -> test  [label="Design new experiment"];
  test -> evaluation  [label="Statistical analysis"];
  evaluation -> data  [label="Knowledge accumulation"];

>>>>

The process described here can be largely automated in R using the `theorytools` package; see the package vignette on FAIR theory, `vignette("fairtheory", package = "theorytools")`.

@@

```{r include = FALSE}
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)
ec_groot <- grViz(diagram = 
'digraph neato { node [fontsize = 6]; 
    edge [fontsize = 6];
    overlap = false; fontsize = 7;
  graph [layout = neato]

  observation [pos = "2,3!"];
  induction [pos = "3,2!"];
  deduction [pos = "3,1!"];
  test [pos = "1,1!"];
  evaluation [pos = "1,2!"];

}')
ec_wagenmakers <- grViz(diagram = 
'digraph neato { node [fontsize = 6]; 
    edge [fontsize = 6];
    overlap = false; fontsize = 7;
  graph [layout = neato]

  subgraph cluster_discovery {
    label="Discovery";
    induction [label="New hypothesis" pos = "3,2!"];
    deduction [label="New prediction" pos = "3,1!"];
  }
  observation  [label="Old knowledge and old data" pos = "2,3!"];
  subgraph cluster_justification {
    label="Justification";
    test [label="Test on new data" pos = "1,1!"];
    evaluation [pos = "1,2!"];
  }

  observation -> induction [label="Speculate & explore" labeljust=r];
  induction -> deduction  [label="Deduce"];
  deduction -> test  [label="Design new experiment"];
  test -> evaluation  [label="Statistical analysis"];
  evaluation -> observation  [label="Knowledge accumulation"];

}')
rsvg_svg(charToRaw(export_svg(ec_groot)), "ec_groot.svg")
rsvg_svg(charToRaw(export_svg(ec_wagenmakers)), "graph.svg")
```

@@

## Automating these Steps

R-users can use the `theorytools` package to partly automate the preceding steps, for example, using following code (see the package documentation for more information):

```
install.packages("theorytools")
library(theorytools)
# Use worcs to check if GitHub permissions are set:
library(worcs)
check_git()
check_github()
# Create the theory repository:
fair_theory(path = "c:/theoryfolder/empirical_cycle",
            title = "The Empirical Cycle",
            theory_file = "empirical_cycle.dot",
            remote_repo = "empirical_cycle",
            add_license = "cc0")
```

Note that this function also automatically provides basic FAIR theory metadata to Zenodo.

## Forking Different Implementations of a Theory

De Groot's empirical cycle has inspired several authors,
but closer examination reveals that not all of them have interpreted his work in the same way.
For example, Wagenmakers and colleagues [-@wagenmakersCreativityVerificationCyclePsychological2018] write *"De Groot’s “empirical cycle,” shown here in Figure 6"* - but Figure 6 diverges substantially from De Groot's description, and from our implementation of it.
One advantage of FAIR theory is that we can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
This illustrates an important advantage of FAIR theory:
we can implement different versions of a theory, and study their relationships.
In practice, this is achieved by ["forking the repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo),
["cloning"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo) the forked repository to our local computer,
making any changes we want, and then completing steps 4-5 of "Making a Theory FAIR".

We attempted to implement Wagenmakers and colleagues' version as a DOT graph,
to illustrate some clear deviations from the original.
The authors deviate from De Groot in a minor way by renaming the five phases. 
While this change was not made explicit (thus illustrating an advantage of FAIR theory), we assumed that the phases are identical, just with different labels.
The authors do mention one explicit change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*.
The DOT graph below illustrates this change:

@@

  subgraph cluster_discovery {
    label="Discovery";
    induction [label="New hypothesis"];
    deduction [label="New prediction"];
  }
  observation  [label="Old knowledge and old data"];
  subgraph cluster_justification {
    label="Justification";
    test [label="Test on new data"];
    evaluation;
  }

  observation -> induction [label="Speculate & explore"];
  induction -> deduction  [label="Deduce"];
  deduction -> test  [label="Design new experiment"];
  test -> evaluation  [label="Statistical analysis"];
  evaluation -> observation  [label="Knowledge accumulation"];

### 2025-01-02 manuscript.tex
add svg code for DOT graphs

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 6763}
\usepackage{lineno}

@@

This is a preprint paper, generated from Git Commit \# 70a3322a.

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' \emph{empirical content} {[}REF{]} as it forces researchers to use precise statements, for example by specifying exact functional forms for relations or forces them to specify processes that would otherwise get away with their vague formulation in verbal theories.
For example, the phonological loop in Baddeley's verbal description of his working memory model allows at least 144 different implementations, for example, depending on how decay rate, recall success, or rehearsal sequence are precisely implemented (Lewandowsky \& Farrell, 2010).
More precise theories makes them easier to falsify,
which necessitates revising them,
thus advancing our principled understanding of the phenomena they describe.
FAIR theory does not require theories to be formal,
and formal theory can be represented in a way that is not FAIR.
It is therefore important for us to emphasize that FAIR theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down.
The guidelines introduced by FAIR theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
Conversely, a directed acyclic graph (DAG) is a type of formal theory,
but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
then this formal theory is not FAIR.
FAIR theory is thus consistent with, but does not require, formal theory (also see \emph{Accessibility}).

@@

Open science infrastructure is an area of active development, and as such,
the approach proposed here should not be considered definitive,
but rather, as one proposal for a FAIR-compliant implementation of theory using infrastructure available at the time of writing.

--\textgreater{}
--\textgreater{}

@@

Create a new folder and copy the theory file from the previous step into it.
To help meet the Interoperability and Reusability criteria,
add two more files:
A README.md file with instructions for future users of your theory,
and a LICENSE file with the legal conditions for reuse.
We recommend using the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.

@@

\begin{itemize}
\tightlist
\item
  Title, prefaced with \texttt{\#\ FAIR\ theory:\ The\ Theory\textquotesingle{}s\ Name}
\item
  Description: A plain-text description of the theory and its scope
\item
  Interoperability: This section, also often labeled ``Getting Started'', ``Instructions'', or ``How to Use'' might be better labeled Interoperability from a FAIR perspective. Following the discussion in the respective section, we propose focusing on X-interoperability, and telling users exactly what they can use the theory for, and how. For example, our example is implemented in the DOT language for describing graphs, so we would provide instructions here on how to plot a DOT graph.
\item
  Contributing: Pertaining to the Reusability criterion, this section should tell users the \emph{social expectations regarding reuse and contributions}.
\item
  License: As a complement to the preceding section, this section should refer readers to the LICENSE file to learn about the \emph{legal constraints to reuse}.
\item
  Citing this work: Tell users how to cite the theory.
\item
  Related works: This section should refer to the work that the FAIR theory is derived from, or documented in. Note that this is redundant with metadata entered in Zenodo (step 5). The advantage is that documenting related works in the README makes it more readily accessible to users. The disadvantage is that you may have to maintain this section of the readme. A compromise might be to use this section to document one canonical reference for the theory that is unlikely to change going forward. For example, we referenced the original empirical cycle paper here:
\end{itemize}

\begin{verbatim}
This repository contains an implementation of the "empirical cycle",
a model proposed by De Groot and Spiekerman (1969, p. 28)

@@

The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
In particular, the practice of version control offers extensive benefits for scientific work Van Lissa et al. (2021).
To version control our project, we initiate a Git repository in the project folder.
This can be done by \href{https://git-scm.com/book/en/v2/Getting-Started-Installing-Git}{installing Git} locally,
opening a terminal in the project folder,
and running the command \texttt{git\ init}.
We subsequently create a remote repository to host a copy of this local Git repository, which will in turn be archived.
To \href{https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/}{do this on GitHub}, create an account and press the button labeled \texttt{Create\ new\ repository}.
The default settings for the repository are fine;
for the next steps of the tutorial it must be set to ``Public''.
Copy the repository's URL, which usually has the form \texttt{https://github.com/username/repository.git}.
Next, in the terminal window, run the following lines of code sequentially:

\begin{verbatim}
git add .
git commit -m "First commit"
git remote add origin *[the repository URL you copied]*
git branch -M main
git push -u origin main
\end{verbatim}

The local files should now be backed up to the Git remote repository;
refresh the webpage to verify.

@@

The process of archiving a GitHub repository on Zenodo is documented in a vignette in the \texttt{theorytools} R-package, so that it can be kept up-to-date.
We present a brief summary of the instructions at the time of writing here.
First, create a Zenodo account with your existing GitHub account.
Then in Zenodo, go to the GitHub section under your account.
Syncronize Zenodo to import all of your public GitHub repositories.
Following the instructions on the page, activate Zenodo for your theory repository.
Then, head to the repository's GitHub page, and create a new release.
You have to choose a tag and release title;
we suggest using semantic versioning for both, starting with version \texttt{0.1.0}, unless you use another convention for versioning (which should be documented in README.md).
After publishing the release,
you should be able to see it in your Zenodo account.

@@

\begin{itemize}
\tightlist
\item
  Set the \emph{resource type} to \texttt{Model}
\item
  Verify that the \emph{title} is prefaced with \texttt{FAIR\ theory:}
\item
  Add the \emph{keyword} \texttt{fairtheory}
\item
  List the DOIs/identifiers of \emph{related works}. Use the \texttt{Relation} field as appropriate. For example:

@@

@@

```
digraph \{

theory;
prediction;
test {[}label=``inferential procedure''{]};
observation;

theory -\textgreater{} prediction {[}label=``deduction''{]};
prediction -\textgreater{} test;
test -\textgreater{} observation;
observation -\textgreater{} theory {[}label=``generalization''{]};

\}

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 6897}
\usepackage{lineno}

@@

This is a preprint paper, generated from Git Commit \# 9c5837f3.

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' \emph{empirical content} {[}REF{]} as it forces researchers to use precise statements, for example by specifying exact functional forms for relations or forces them to specify processes that would otherwise get away with their vague formulation in verbal theories.
For example, the phonological loop in Baddeley's verbal description of his working memory model allows at least 144 different implementations, for example, depending on how decay rate, recall success, or rehearsal sequence are precisely implemented (Lewandowsky \& Farrell, 2010).
More precise theories makes them easier to falsify,
which necessitates revising them,
thus advancing our principled understanding of the phenomena they describe.
FAIR theory does not require theories to be formal,
and formal theory can be represented in a way that is not FAIR.
It is therefore important for us to emphasize that FAIR theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down.
The guidelines introduced by FAIR theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
Conversely, a directed acyclic graph (DAG) is a type of formal theory (for example, see the empirical cycle in Figure \ref{fig:figec}),
but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
it is not FAIR.
FAIR theory is thus consistent with, but does not require, formal theory (also see \emph{Accessibility}).

@@

Open science infrastructure is an area of active development, and as such,
the approach proposed here should not be considered definitive,
but rather, as one proposal for a FAIR-compliant implementation of theory.
At the time of writing (2024), the integration of GitHub and Zenodo makes for a particularly user-friendly approach.
Nevertheless, it is important to stress that alternatives to GitHub (Gitlab, Bitbucket, etc.) and Zenodo (e.g., institutional repositories) exist.
The principles described here
(using version control and archiving major versions) could be implemented in a different workflow.

The process described here can be largely automated in R using the \texttt{theorytools} package; see the package vignette on FAIR theory, \texttt{vignette("fairtheory",\ package\ =\ "theorytools")}.

@@

Create a new folder and copy the theory file from the previous step into it.
To help meet the Interoperability and Reusability criteria,
add two more files:
A README.md file with instructions for future users of your theory,
and a LICENSE file with the legal conditions for reuse.
We recommend the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.

@@

\begin{itemize}
\tightlist
\item
  Title, prefaced with \texttt{\#\ FAIR\ theory:\ The\ Theory\textquotesingle{}s\ Name}
\item
  Description: A plain-text description of the theory and its scope
\item
  Interoperability: Most README files contain a section labeled ``Getting Started'', ``Instructions'', or ``How to Use''. From a FAIR perspective, such a section might be better labeled ``Interoperability'', or ``How to Use (Interoperability)''. We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how. For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
\item
  Contributing: Pertaining to the Reusability criterion, this section should tell users the \emph{social expectations regarding reuse and contributions}.
\item
  License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the \emph{legal conditions of reuse}.
\item
  Citing this work: Tell users how to cite the theory. Note that this section is redundant with the Zenodo archive, which has a preferred citation field. The disadvantage of redundant information is that you may have to maintain this section of the README going forward. The advantage is that documenting related works in the README makes it more readily accessible to users. We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
\item
  Related works: This section should refer to the work that the FAIR theory is derived from, or documented in. Again, this is redundant with metadata entered in Zenodo (step 5). We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward. For example, we referenced the original empirical cycle paper here:
\end{itemize}

\begin{verbatim}
This repository contains an implementation of the "empirical cycle",
a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.

@@

The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
In particular, the practice of version control offers extensive benefits for scientific work Van Lissa et al. (2021).
To version control our project, we initiate a Git repository in the project folder.
We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
Note that the repository
must be set to ``Public'' to take advantage of GitHub's Zenodo integration.

Push the local files to the Git remote repository, and keep them synchronized going forward.

@@

The process of archiving a GitHub repository on Zenodo is documented in a vignette in the \texttt{theorytools} R-package, so that it can be kept up-to-date.
We present a brief summary of the instructions at the time of writing here.
First, create a Zenodo account with your existing GitHub account.
Then in Zenodo, go to the GitHub section under your account.
Following the instructions on the page, activate Zenodo for your theory repository.
Then, create a new release of the GitHub repository.
You have to choose a tag and release title;
we suggest using semantic versioning for both, starting with version \texttt{0.1.0}, unless you use another convention for versioning (which should be documented in README.md).
After publishing the release,
you should be able to see the archived version in your Zenodo account,
along with a DOI.

@@

\begin{itemize}
\tightlist
\item
  Set the \emph{resource type} to \texttt{Model}
\item
  Verify that the \emph{title} is prefaced with \texttt{FAIR\ theory:}
\item
  Add the \emph{keyword} \texttt{fairtheory}
\item
  Optionally, submit the theory to the \href{https://zenodo.org/communities/fairtheory}{``FAIR Theory Community''} to increase its findability
\item
  List the DOIs/identifiers of \emph{related works}. Use the \texttt{Relation} field as appropriate. For example:

@@

\subsection{Automating these Steps}\label{automating-these-steps}

R-users can use the \texttt{theorytools} package to partly automate the preceding steps, for example, using following code (see the package documentation for more information):

\begin{verbatim}
install.packages("theorytools")
library(theorytools)
# Use worcs to check if GitHub permissions are set:
library(worcs)
check_git()
check_github()
# Create the theory repository:
fair_theory(path = "c:/theoryfolder/empirical_cycle",
            title = "The Empirical Cycle",
            theory_file = "empirical_cycle.dot",
            remote_repo = "empirical_cycle",
            add_license = "cc0")
\end{verbatim}

Note that this function also automatically provides basic FAIR theory metadata to Zenodo.

@@

  theory;
  prediction;
  test [label="inferential procedure"];
  observation;

  theory -> prediction [label="deduction"];
  prediction -> test;
  test -> observation;
  observation -> theory [label="generalization"];
